---
title: "Tree management practices and outcomes"
author: "Cory Whitney"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

library(bibtex)
library(tm)
library(tidyverse)

```

Assess associations between tree management practices and outcomes. Start with term frequency data and bibliographic entries. 

```{r assess_refs, warning = FALSE, message=FALSE}

# Load BibTeX data
bib_data <- read.bib("refs/scopus.bib")
bib_keys <- sapply(bib_data, function(x) x$key)

# Extract text (title + abstract)
bib_df <- data.frame(
  key = bib_keys,
  text = sapply(bib_data, function(x) paste(x$title, x$abstract, x$keywords, sep = " "))
)
```

Preprocess the text corpus

```{r pre_process_bib_data, warning = FALSE, message=FALSE}

corpus <- VCorpus(VectorSource(bib_df$text))
corpus <- corpus %>%
  tm_map(content_transformer(tolower)) %>%
  tm_map(removePunctuation) %>%
  tm_map(removeNumbers) %>%
  tm_map(removeWords, stopwords("en")) %>%
  # stemming to reduce words to their base / root form
  tm_map(stemDocument) %>%  # Add stemming
  tm_map(stripWhitespace)

```

Check for dropped papers

```{r check_dropped_documents, warning = FALSE, message=FALSE}

source("functions/log_dropped_docs.R")

corpus <- log_dropped_docs(corpus, content_transformer(tolower), "Convert to lowercase")
corpus <- log_dropped_docs(corpus, removePunctuation, "Remove punctuation")
corpus <- log_dropped_docs(corpus, removeNumbers, "Remove numbers")
# corpus <- log_dropped_docs(corpus, removeWords, stopwords("en"), "Remove stopwords")
corpus <- log_dropped_docs(corpus, stemDocument, "Apply stemming")

```

There were 7,876 returns in the Scopus search of Title, Abstract and Keywords: `( "agroforestry" OR "silviculture" OR "conservation forest" OR "community forest" ) + ( "effect" OR "development" OR "improvement" OR "potential" ) + ( "management" )`. 

Compiling and remove duplicates and erroneous returns for a total of `r length(corpus)` papers in search. Use these to create a Term-Document Matrix (TDM). 

```{r create_tdm, warning = FALSE, message=FALSE}
#TDM
tdm <- TermDocumentMatrix(corpus)

# Convert TDM to a matrix 
tdm_matrix <- as.matrix(tdm)

```

Create an inspection process for frequently occurring terms. 

```{r frequently_occurring_terms}
# Inspect frequently occurring terms
term_freq <- rowSums(tdm_matrix)
term_freq_sorted <- sort(term_freq, decreasing = TRUE)
head(term_freq_sorted, 50)

```

Create an overview of the associations between tree management and relevant outcomes from the references. 

```{r define_practices_outcomes, warning = FALSE, message=FALSE}
source("functions/workflow_tkik_outcomes.R")
# takes a while to run
```

Create an overview of practice keywords in the reference search. 

```{r check_keywords}
# Check keyword presence (check how often practices show up in the literature)
practice_check <- sapply(practices, function(practice) {
  sum(grepl(practice, text_data, ignore.case = TRUE))
})
print(practice_check)
```

Overview the results of the outcome keywords in the reference search. 

```{r check_outcomes, warning = FALSE, message=FALSE}
# (check how often outcomes show up in the literature)
outcome_check <- sapply(outcomes, function(outcome) {
  sum(grepl(outcome, text_data, ignore.case = TRUE))
})
print(outcome_check)
```

Ready the reference data for a plot of associations. 

```{r arrange_plot_associations, warning = FALSE, message=FALSE}
library(ggplot2)
library(stringr)

# Create human-readable labels by replacing underscores or capitalization patterns
association_df$PracticeGroup <- str_replace_all(association_df$PracticeGroup, "_", " ")
association_df$PracticeGroup <- str_to_title(association_df$PracticeGroup) # Convert to title case

association_df$OutcomeGroup <- str_replace_all(association_df$OutcomeGroup, "_", " ")
association_df$OutcomeGroup <- str_to_title(association_df$OutcomeGroup) # Convert to title case

```

Create an `OutcomeGroup` and check that the values match categories. 

```{r group_outcomes, warning = FALSE, message=FALSE}
# Define the outcome categories
human_development_outcomes <- c("Economic", "Livelihoods", "Nutrition")
biodiversity_outcomes <- c("Carbon", "Biodiversity", "Ecosystem Services")

# Add the classification column for outcomes
association_df$OutcomeCategory <- ifelse(
  association_df$OutcomeGroup %in% human_development_outcomes, 
  "Human Development",
  ifelse(
    association_df$OutcomeGroup %in% biodiversity_outcomes,
    "Biodiversity",
    NA  # Assign NA if it doesn't match any category
  )
)

# Check for any NA values in OutcomeCategory
if (any(is.na(association_df$OutcomeCategory))) {
  warning("Some OutcomeGroup values did not match any category. Please review the data.")
}
```

Plot of the associations between tree management and outcomes for people and the environment. 

```{r order_x-axis}
# Ensure proper ordering of x-axis labels grouped by category
all_outcomes <- c(human_development_outcomes, biodiversity_outcomes)

association_df$OutcomeGroup <- factor(
  association_df$OutcomeGroup,
  levels = all_outcomes
)
```

```{r plot_grouped}
# Bubble plot with grouped x-axis
association_plot <- ggplot(association_df, aes(x = OutcomeGroup, y = PracticeGroup, size = Count)) +
  geom_point(shape = 21, color = "black", alpha = 0.7) + 
  scale_size_continuous(name = "Count", range = c(2, 10)) +
  labs(
    title = "Associations Between Forest Practices and Outcomes",
    x = "Outcome",
    y = "Forest Practice"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_line(color = "gray80", linetype = "dotted"),
    panel.grid.minor = element_blank()
  ) +
  facet_grid(~ OutcomeCategory, scales = "free_x", space = "free")

association_plot

ggsave(filename = "figures/association_plot.png", plot = association_plot, width = 5, height = 5, bg = "white")
```


