---
title: "Tree management practices and outcomes"
author: "Cory Whitney"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

```

Assess associations between tree management practices and outcomes. Start with term frequency data and bibliographic entries. 

```{r assess_refs}
library(NLP)
source("functions/create_corpus.R")
```

We found a total of `r length(corpus)` papers in the Scopus search. We use these to create a Term-Document Matrix (TDM). 

```{r create_tdm}
#TDM
tdm <- TermDocumentMatrix(corpus)

# Convert TDM to a matrix 
tdm_matrix <- as.matrix(tdm)

# Save the TDM matrix to a file
write.csv(tdm_matrix, "data/term_document_matrix.csv", row.names = TRUE)
```

Create an inspection process for frequently occurring terms. 

```{r frequently_occurring_terms}
# Inspect frequently occurring terms
term_freq <- rowSums(tdm_matrix)
term_freq_sorted <- sort(term_freq, decreasing = TRUE)
head(term_freq_sorted, 20)

```

Create an overview of the associations between tree management and relevant outcomes from the references. 

```{r define_practices_outcomes}
source("functions/make_association_dataframe.R")

# Save the associations to a CSV file
write.csv(association_df, "forest_practice_outcome_associations.csv", row.names = FALSE)
```

A total of `r sum(association_df$Count)` references had some mention of the relevant associations. 

Create an overview the keywords of the outcomes in the reference search. 

```{r check_keywords}
# Check keyword presence
keyword_check <- sapply(practices, function(practice) {
  sum(grepl(practice, text_data, ignore.case = TRUE))
})
print(keyword_check)
```
Overiew the results of the outcomes in the reference search. 

```{r check_outcomes}
outcome_check <- sapply(outcomes, function(outcome) {
  sum(grepl(outcome, text_data, ignore.case = TRUE))
})
print(outcome_check)
```
Ready the reference data for a plot of associations. 

```{r arrange_plot_associations}
library(ggplot2)
library(stringr)

# Create human-readable labels by replacing underscores or capitalization patterns
association_df$PracticeGroup <- str_replace_all(association_df$PracticeGroup, "_", " ")
association_df$PracticeGroup <- str_to_title(association_df$PracticeGroup) # Convert to title case

association_df$OutcomeGroup <- str_replace_all(association_df$OutcomeGroup, "_", " ")
association_df$OutcomeGroup <- str_to_title(association_df$OutcomeGroup) # Convert to title case

```

Create an `OutcomeGroup` and check that the values match categories. 

```{r group_outcomes}
# Define the outcome categories
human_development_outcomes <- c("Economic", "Livelihoods", "Nutrition")
biodiversity_outcomes <- c("Carbon", "Biodiversity", "Ecosystem Services")

# Add the classification column for outcomes
association_df$OutcomeCategory <- ifelse(
  association_df$OutcomeGroup %in% human_development_outcomes, 
  "Human Development",
  ifelse(
    association_df$OutcomeGroup %in% biodiversity_outcomes,
    "Biodiversity",
    NA  # Assign NA if it doesn't match any category
  )
)

# Check for any NA values in OutcomeCategory
if (any(is.na(association_df$OutcomeCategory))) {
  warning("Some OutcomeGroup values did not match any category. Please review the data.")
}
```

Plot of the associations between tree management and outcomes for people and the environment. 

```{r order_x-axis}
# Ensure proper ordering of x-axis labels grouped by category
all_outcomes <- c(human_development_outcomes, biodiversity_outcomes)

association_df$OutcomeGroup <- factor(
  association_df$OutcomeGroup,
  levels = all_outcomes
)
```

```{r plot_grouped}
# Bubble plot with grouped x-axis
ggplot(association_df, aes(x = OutcomeGroup, y = PracticeGroup, size = Count, fill = Count)) +
  geom_point(shape = 21, color = "black", alpha = 0.7) +
  scale_fill_gradient(low = "white", high = "blue", name = "Count") +
  scale_size_continuous(name = "Count", range = c(2, 10)) +
  labs(
    title = "Associations Between Forest Practices and Outcomes",
    x = "Outcome",
    y = "Forest Practice"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_line(color = "gray80", linetype = "dotted"),
    panel.grid.minor = element_blank()
  ) +
  facet_grid(~ OutcomeCategory, scales = "free_x", space = "free")
```


